### 2015-05-15 background.tex
Removed subsections from background and replaced with paragraph headings.  A bit reluctant to do this, but I think it makes sense to save cutting too much text.
Pulled background back a page.
Done it.
fix

\subsection{Concurrency in Device Drivers}
\label{bg:concurrency}

Modern operating systems address the demand for responsiveness and performance in device drivers by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.  These forms of concurrent execution are prone to \emph{data races}.

@@

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux driver. The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} object as a parameter. However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux driver}
\label{fig:data_race_example}
\end{figure}

The most common method for avoiding races is by protecting sets of program statements that access a shared resource using \emph{locks}, forming \emph{critical sections}.  Figure~\ref{fig:lock_example} shows how to use a lock to eliminate the races in Figure~\ref{fig:data_race_example}. Because the return statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency; while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\end{figure}

\subsection{Lockset Analysis}
\label{bg:lockset}

Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.  The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a \emph{potential} race on that memory location. This is because an empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
\end{figure}

>>>>

\noindent\textbf{Concurrency in Device Drivers }
%
Modern operating systems address the demand for responsiveness and performance in device drivers by providing multiple sources of concurrency~\cite{corbet2005linux}: the kernel can invoke an arbitrary number of entry points from the same driver concurrently; a running driver process can block and be replaced by another process in the same driver; and hardware interrupts can be handled concurrently.  These forms of concurrent execution are prone to \emph{data races}.

@@

Figure~\ref{fig:data_race_example} shows a racy entry point, \texttt{nvram\_llseek}, in the generic\_nvram Linux driver. The driver can invoke the entry point concurrently from two threads $T_1$ and $T_2$, with the same \texttt{file} struct as a parameter. However, this can lead to multiple possible data races, because $T_1$ and $T_2$ can access the \texttt{f\_pos} field of \texttt{file} in arbitrary order. Our tool, \whoop (see \S\ref{whoop}), was able to find these races automatically (see \S\ref{evaluation}).

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break; // racy
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset; // racy
  return file->f_pos; // racy
}
\end{lstlisting}
\vspace{-2mm}
\caption{Racy entry point in the generic\_nvram Linux driver}
\label{fig:data_race_example}
\vspace{-2mm}
\end{figure}

The most common method for avoiding races is by protecting sets of program statements that access a shared resource with \emph{locks}, forming \emph{critical sections}.  Figure~\ref{fig:lock_example} shows how to use locking to eliminate the races in Figure~\ref{fig:data_race_example}. Because the \texttt{return} statement can potentially race on the \texttt{f\_pos} field, we store the result in a temporary variable inside the critical section.
%
Careless use of locks has many well-known pitfalls~\cite{sutter2005software}: coarse-grained locking can hurt performance, as it limits the opportunity for concurrency; while fine-grained locking can easily lead to deadlocks. Although the Linux kernel provides sophisticated lock-free synchronization mechanisms~\cite[p.\ 123]{corbet2005linux}, such as atomic read-modify-write operations, in this work we focus on locks as they are widely used.\footnote{We treat lock-free operations soundly, by not regarding them as providing any protection between threads, but this can lead to false alarm race reports.}

\begin{figure}[t]
\begin{lstlisting}
static loff_t nvram_llseek(struct file *file,
    loff_t offset, int origin) {
  mutex_lock(&nvram_mutex); // lock
  switch (origin) {
    case 0: break;
    case 1: offset += file->f_pos; break;
    case 2: offset += nvram_len; break;
    default: offset = -1;
  }
  if (offset < 0)
    return -EINVAL;
  file->f_pos = offset;
  loff_t res = file->f_pos; // store result
  mutex_unlock(&nvram_mutex); // unlock
  return res;
}
\end{lstlisting}
\vspace{-2mm}
\caption{Using a lock to eliminate the data races in the previous example}
\label{fig:lock_example}
\vspace{-2mm}
\end{figure}

\noindent\textbf{Lockset Analysis }
%
Lockset analysis is a lightweight race detection method proposed in the context of Eraser~\cite{savage1997eraser}, a dynamic data race detector.  The idea is to track the set of locks that are \emph{consistently} used to protect a memory location during program execution. If that lockset ever becomes empty, the analysis reports a \emph{potential} race on that memory location. This is because an empty lockset suggests that a memory location \emph{may} be accessed simultaneously by two or more threads.

@@

\begin{figure}[t]
\centering
\includegraphics[width=1\linewidth]{img/lockset.pdf}
\vspace{-2mm}
\caption{Applying lockset analysis on a concurrent program}
\label{fig:locksets}
\vspace{-2mm}
\end{figure}
